---
title: 'Final Project Code'
author: 'Rebecca Baumann, Chang Gao, Jeffery Qian'
# output: html_document
# embed-resources: true
format: html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,tidy.opts = list(width.cutoff = 60), tidy = TRUE)
```
Model A:
```{r}
library(deSolve)
```

Part 1: Equations 
```{r}
# Parameters
r_R <- 0.5      # resident whale growth rate
K_R <- 30       # resident carrying capacity

r_P <- 0.8      # pinniped growth rate
K_P <- 500      # pinniped carrying capacity

alpha <- 0.2    # effect of pinnipeds on residents
beta  <- 0.6    # effect of residents on pinnipeds

C   <- 0.02     # predation rate
B   <- 0.2      # conversion efficiency
D_T <- 0.1      # transient whale death rate
```

Initialization
```{r}
R0 <- 10
P0 <- 200
T0 <- 5

state_ini <- c(R = R0,
               P = P0,
               T = T0)
```

Model A equations
```{r}
modelA <- function(time, state, parameters){

  R <- state[1]
  P <- state[2]
  T <- state[3]

  # Resident whales (competition with P)
  dR <- r_R * R * (1 - (R + alpha*P)/K_R)

  # Pinnipeds (competition + predation)
  dP <- r_P * P * (1 - (P + beta*R)/K_P) -
        C * P * T

  # Transient whales (predator)
  dT <- B * C * P * T -
        D_T * T

  return(list(c(dR, dP, dT)))
}
```

Add time to this function
```{r}
time_vec <- seq(0, 200, length.out = 2000)

outA <- ode(y = state_ini,
            times = time_vec,
            func = modelA,
            parms = NULL)

outA <- as.data.frame(outA)
```

Plot
```{r}
plot(outA$time, outA$R,
     type = "l",
     col = "blue",
     lwd = 2,
     ylim = c(0,600),
     xlab = "Time",
     ylab = "Population")

lines(outA$time, outA$P,
      col = "darkgreen",
      lwd = 2)

lines(outA$time, outA$T,
      col = "red",
      lwd = 2)

legend("topright",
       legend = c("Residents (R)",
                  "Pinnipeds (P)",
                  "Transients (T)"),
       col = c("blue","darkgreen","red"),
       lty = 1,
       lwd = 2)
```
Conclusion: This graph proved that the presence of residents decreases the equilibrium population size of transients and transient increases residents as discussed in the paper. The system exhibits damped oscillations before converging to a stable interior equilibrium.
The transient whales suppress pinnipeds, which reduces competition pressure on resident whales, allowing residents to increase in density.

Part 2:Equilibrium 
```{r}
# Equilibrium (Model A)
P_star <- D_T / (B * C)

R_star <- K_R - (alpha * D_T) / (B * C)

T_star <- (r_P / (C * K_P)) *
          (K_P - (D_T / (B * C)) -
           beta * K_R +
           (alpha * beta * D_T) / (B * C))

P_star
R_star
T_star
```

```{r}
dR = 0
dP = 0
dT = 0

# Test derivatives at equilibrium
dR_test <- r_R * R_star *
           (1 - (R_star + alpha * P_star) / K_R)

dP_test <- r_P * P_star *
           (1 - (P_star + beta * R_star) / K_P) -
           C * P_star * T_star

dT_test <- B * C * P_star * T_star -
           D_T * T_star

dR_test
dP_test
dT_test

```

Part 3: Equilibrium when absence
```{r}
# Equilibrium without competition
R_not <- (K_R - alpha * K_P) /
         (1 - alpha * beta)

T_not <- (r_P / (C * K_P)) *
         (K_P - (D_T / (B * C)))

R_not
T_not
R_star > R_not
T_star < T_not
```
Since both are true, plus-minus interaction is also true.

Part 4: Test if the transient whale population will be smaller when both whales are present
```{r}
B * C * K_R > alpha * D_T
```

This is correspondence with the paper's conclusion.


Model B: resident whales and pinnipeds have an indirect interaction.
$$
% \begin{aligned}
% R'=&R(B_RC_RS-D_R) &\rm{resident\ whales}\\
% T'=&T(B_TC_TP-D_T)&\rm{transient\ whales}\\
% P'=&r_PP(1-\frac{P}{K_P}-\frac{\alpha S}{K_P})-C_TPT&\rm{pinnipeds}\\
% S'=&r_SS(1-\frac{S}{K_S}-\frac{\beta P}{K_S})-C_RSR &\rm{salmon}
% \end{aligned}
\begin{aligned}
R'=&R(B_RC_RS-D_R)\\
T'=&T(B_TC_TP-D_T)\\
P'=&r_PP(1-\frac{P}{K_P}-\frac{\alpha S}{K_P})-C_TPT\\
S'=&r_SS(1-\frac{S}{K_S}-\frac{\beta P}{K_S})-C_RSR
\end{aligned}
$$
Observation: natural death rate of pennipeds and salmon seem to be ignored.

Parameters:
```{r}
B_R=1 # birth rate of resident whales
C_R=0.002 # Consumption/conversion rate of resident whales
D_R=1 # death rate of resident whales

B_T=1 # birth rate of transient whales
C_T=0.002 # Consumption/conversion rate of resident whales
D_T=1 # death rate of transient whales

r_P=30 # growth rate of pinnipeds
K_P=1000 # Carrying capacity of pinnipeds
a=1 # attack rate on pinnipeds

r_S=30 # growth rate of salmon
K_S=1000 # Carrying capacity of salmon
b=1 # attack rate on salmon
```

```{r}
library(deSolve)
require(deSolve)

# predator-prey model
LV <- function(time,state,para) {
	B_R=para[1];C_r=para[2];D_R=para[3];B_T=para[4];C_T=para[5];D_T=para[6];r_P=para[7];K_P=para[8];a=para[9];r_S=para[10];K_S=para[11];b=para[12]
	# state variables
	R=state[1];T=state[2];P=state[3];S=state[4]
	# equation
	X1=R*(B_R*C_R*S-D_R)
	X2=T*(B_T*C_T*P-D_T)
	X3=r_P*P*(1-P/K_P-a*S/K_P)-C_T*P*T
	X4=r_S*S*(1-S/K_S-b*P/K_S)-C_R*S*R

	return(list(c(X1,X2,X3,X4)))
}

# input parameters
para=c(B_R,C_R,D_R,B_T,C_T,D_T,r_P,K_P,a,r_S,K_S,b)
# initialization
time=seq(from=0,to=100,length=1000)
ini=c(1,1,30,100)
out=ode(ini,time,LV,para)
```
Plot
```{r}
plot(out,ylab="Population")
```
It seems that the population for resident whales and transient whales almost periodically dies out every 50 years. Pinnipeds and salmon population exhibit periodic change in this parameter setting.