---
title: 'Code testing region'
author: 'name'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 3 main populations:
 P: Pinnipeds
 R: Resident Whales
 T: Transient Whales

# 2 models: 1 is direct competition, 2 includes indirect competition

# Part 1: Direct Competition Model

#Basic interactions for competition represented by:
 dP/dt = r_P*P[(K_P-P-beta*R)/K*P] - C*P*T
 dR/dt = r_R*R(K_R-R-alpha*P)/K_R
 dT/dt = T*(B*C*P-D_T)
 
 
r_R: intrinsic rate of increase (resident whales)
K_R: carrying capacity (resident whales)
r_P:intrinsic rate of increase (pinnipeds)
K_P: carrying capacity (pinnipeds)
alpha: competition coefficient between pinnipeds and resident whales
beta: competition coefficient between resident whales and pinnipeds
C: the number of pinnipeds captured per unit of time per unit pinniped density by an average transient whale (basically consumption rate)
B: the efficiency by which transient whales consume and assimilate pinnipeds
D_T: the density-independent death rate of transient whales

Notes: beta should be much larger than alpha
K_R should be much smaller than K_P
product of the competition coefficients should be less than 1

```{r}
# note: I am assigning these ones made up values, although I am following guidelines of x must be bigger than y, z*a must be greater than/less than zero, etc. generally just not sure what should be put
r_R <-0.5
r_P <-0.75
K_R <- 30
K_P <- 500
alpha <- 5
beta <- 150


C <- 5 #assuming days time unit for now, like all whales catch 5/day style parameter
B <- 0.1 #(based off of the 10% rule for now)
D_T <- 0.04 #unsure what to put here, this is a test value
tsetORCA <- seq(from = 0, to = 100, length.out = 20000)
R.simuset <- NaN*tsetORCA
```


```{r}
plot(x = tsetORCA, y = R,
     type = 'l', las = 1, lwd = 2, col = Xcol,
     xlab = 'Time', ylab = 'Resident whales')
# fix plot values

```


#Equations have a single equilibrium point with positive population densities of all three types. This is specified by:

P = D_T/(B*C)
R = K_R - (alpha*DT/B*C)
T = (r_P/C*K_P)[K_P - (D_T/B*C) - beta*K_R + (alpha*beta*D_T/B*C)]

```{r}

P <- D_T/(B*C)
P
R <- K_R - (alpha*D_T/B*C)
R
T <- (r_P/C*K_P)[K_P - (D_T/B*C) - beta*K_R + (alpha*beta*D_T/B*C)]
T
#uncertain as to why T is orange here
```

#The expressions for equilibrium densities can be compared with the equilibrium density that each type of whale would attain in the absence of the other. These are:

R = (K_R-alpha*K_P)/(1-alpha*beta)
T = (r_P/C*K_P)[Kp-(D_T/B*C)]


```{r}

R2 <- (K_R-alpha*K_P)/(1-alpha*beta)
T2 <- (r_P/C*K_P)[K_P-(D_T/B*C)]

R2
T2

```

#testing the derivative equations

```{r}
#fix errors by setting up dt
 dP/dt <- r_P*P[(K_P-P-beta*R)/K_P] - C*P*T
 dR/dt <- r_R*R(K_R-R-alpha*P)/K_R
 dT/dt <- T*(B*C*P-D_T)

  
```

